---
title: Simulating assemblages from a Gaussian process model
output: pdf_document
---

# Basic setup
```{r}
set.seed(0)
library(assertthat)
library(MASS) # For multivariate normal
```

# Functions

## Making a single kernel

Function that returns a rational quadratic kernel function with the specified mixture of lengthscales.

A rational quadratic kernel has a mixture of lengthscales, with the contribution of high-frequency components being largest with small values of $a$.

```{r}
make_RQ_function = function(l, a){
  # Let's keep l and a the same for all dimensions for now.
  assert_that(length(l) == 1)
  assert_that(length(a) == 1)
  
  function(x){
    # From Rasmussen and Williams
    r = as.matrix(dist(x))
    (1 + r^2 / (2 * a * l^2))^-a
  }
}
```

## Making a multi-species kernel

Function that combines a species-by-species kernel with an environment kernel function (such as the one returned by `make_RQ_function`).

I got the idea for kronecker-structured covariance for multi-species applications from Nick Golding, who got it from James Hensman.

I'm calling the resulting function an eta function because it returns values on the link scale (e.g. logit scale for presence-absence), and thus corresponds to $\eta$ in GLM terminology.

```{r}
make_kronecker_eta = function(species_covariance, kernel_function){
  function(x){
    K = kronecker(species_covariance, kernel_function(x))
    matrix(
      mvrnorm(1, mu = rep(0, nrow(K)), Sigma = K),
      ncol = ncol(species_covariance)
    )
  }
}
```

# A one-dimensional example

Generate an abiotic environment
```{r 1d}
# Generate a multivariate abiotic environment (that happens to fall along a
# single dimension, for easier plotting)
X = seq(0, 4, length = 100) %*% matrix(c(1, 2, .1), nrow = 1)
```

Generate a species covariance matrix
```{r}
# Generate a species covariance matrix that's mostly an identity matrix, but
# with species 1 and 2 (black and red) tightly correlated
n_spp = 4
species_covariance = diag(n_spp)
species_covariance[2, 1] = species_covariance[1, 2] = .9
species_covariance
```

Calculate and plot eta
```{r, cache = TRUE}
# Function to turn environmental observations into eta values
f = make_kronecker_eta(species_covariance, make_RQ_function(l = 2, a = 5))

# Randomly generate eta values and plot them
eta = f(X)
matplot(X[ , 1], eta, type = "l", lty = 1) 
```

Simulate and plot counts from eta
```{r}
counts = matrix(rpois(nrow(X) * n_spp, exp(5 + eta)), ncol = n_spp)
matplot(X[ , 1], counts, log = "y", ylab = "counts (log scale)", cex = 0.5, pch = 1)
```

# A two-dimensional example

```{r 2d, cache = TRUE}
x2 = expand.grid(seq(0, 2, length = 20), seq(0, 5, length = 20))

eta2 = f(x2)

library(ggplot2)
qplot(x2[,1], x2[,2], fill = eta2[, 1]) + geom_raster()
qplot(x2[,1], x2[,2], fill = eta2[, 2]) + geom_raster()
qplot(x2[,1], x2[,2], fill = eta2[, 3]) + geom_raster()
qplot(x2[,1], x2[,2], fill = eta2[, 4]) + geom_raster()
```

